Reproducible Research Project 1 - Easton Bell
==========================================---
output: html_document
---
What is mean total number of steps taken per day?

```{r,echo=FALSE}
## This first line will likely take a few seconds. Be patient!
library (dplyr)
library(reshape)
library(datasets)
library(lattice)
##library(ggplot2)

## load data
setwd("C:/Rtools/")
par(mfrow=c(1,3))
basedata <- read.csv("activity.csv",sep=",",header = TRUE)
#get data set that is devoid of NA's
noNA<-!is.na(basedata["steps"])

noNAsubset<-subset(basedata,noNA,select=c(steps,date))


h<-tapply(noNAsubset$steps,noNAsubset$date,sum)
hcolnames<-c(names(h))

hnoNA<-subset(h,!is.na(h[hcolnames]))

hnoNAcolnames<-c(names(hnoNA))


totalsstepsperday<-sum(hnoNA)/length(hnoNAcolnames)  ## MEAN NUMBER OF STEPS PER DAY
```

```{r,echo=TRUE}
totalsstepsperday
#############################################################################
```
Histogram of the total number of steps taken each day

Report the mean and median of the total number of steps taken per day

What is the average daily activity pattern?

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r, echo=FALSE}
hist(hnoNA,col="blue", main="Steps(NoNA)",xlab="step buckets",ylab="number of days")
abline(v=median(hnoNA),col="magenta")
abline(v=mean(hnoNA),col="red")
##boxplot(hnoNA,col="blue")
```

```{r,echo=TRUE}
mean(hnoNA)

median(hnoNA)
```

```{r,echo=FALSE}

## time series plot of avg steps per day vs interval
###########################################################################################

noNAsubseti<-subset(basedata,noNA,select=c(steps,date,interval))


hi<-melt(noNAsubseti,id=c("date","interval"), measure.vars=c("steps"))
hii<-cast(hi, interval~date, sum)
hcolnamesi<-c(names(hii))
g<-mutate(hii,rowsmm=rowSums(hii))
hiii<-subset(g,select=c(interval,rowsmm))
g1<-mutate(hiii,avgstepsperintervalperday=hiii$rowsmm/(length(hcolnamesi)-1))

maxintv<-subset(g1,g1$avgstepsperintervalperday==max(g1$avgstepsperintervalperday),select=c(interval))
```
```{r,echo=TRUE}
maxintv$interval
```
```{r,echo=FALSE}
plotdata<-g1
d0<-data.frame(intvl=c(g$interval),AvgStepsPerIntvlPerDay=g1$avgstepsperintervalperday)
dx<-d0["intvl"]
dy<-d0["AvgStepsPerIntvlPerDay"]
xaxisname<-"Intervals"
yaxisname<-"Avg Steps per day"
##plot(as.numeric(dx[,1]),as.numeric(dy[,1]), type="l",xaxt="n",main = "Avg Steps Per Intv Per Day",xlab=xaxisname,ylab=yaxisname,ylim=c(min(dy[,1]),max(dy[,1])))


plot(as.numeric(dx[,1]),as.numeric(dy[,1]), type="l",main = "Avg Steps Per Intv Per Day",xlab=xaxisname,ylab=yaxisname,ylim=c(min(dy[,1]),max(dy[,1])))
##axis(1,1:length(dx$intvl),labels=as.char(dx$intvl))

############################################################################################
for(x in 1:length(dx[,1]))
{
  if(dy[x,1]==max(dy[,1]))
    {
    ymax<-dy[x,1]
    xmax<-x   
    xintval<-dx[x,1]
    }
    
}
abline(v=xintval,col="magenta")

##points(xintval, ymax, col = "dark red")


```
Total number of missing values in the dataset (i.e. the total number of rows with NAs)

```{r,echo=FALSE}
isNA<-is.na(basedata["steps"])
countNArows<-sum(isNA) ## number of rows with NA
totrows<-nrow(basedata)## total number of rows
```
number of rows with NA
```{r,echo=TRUE}
sum(isNA)
```
total number of rows
```{r,echo=TRUE}
nrow(basedata)
```

The interval mean value was used to fill in missing values
New total number of steps taken per day. 
```{r,echo=FALSE}
hh<-tapply(basedata$steps,basedata$interval,mean,na.rm=TRUE) ## get subst numbers
hht<-subset(basedata,select=c(steps,interval))

llsteps<-vector()
llinterval<-vector()
lll<-vector()

for (j in 1:length(hht[,1,2]))
    {
        if(!is.na(hht[j,1,2])) 
        {
            lll<-hht[j,,]
            llsteps[j]<-lll$steps
            llinterval[j]<-lll$interval
        }

          else
            { 
              c<-hht[j,,]
              qinterval<-c$interval
              newsteps<-hh[as.character(qinterval)]
              llsteps[j]<-newsteps
              llinterval[j]<-qinterval
            }
        
          
                  
     }
dfclean<-data.frame(steps=llsteps,interval=llinterval)

## new data set with NA substituted for
basedatasubstNA<-mutate(basedata,stepssubstNA=dfclean[,1,2])

hs<-tapply(basedatasubstNA$stepssubstNA,basedatasubstNA$date,sum)
hscolnames<-c(names(hs))


totalsstepsperday<-sum(hs)/length(hs)
```

Histogram of the total number of steps taken each day
Mean and median 

```{r,echo=FALSE}
##################################################################################
hist(hs,col="blue", main="Steps(NA Adjusted For)",xlab="step buckets",ylab="number of days")
abline(v=median(hs),col="magenta")
abline(v=mean(hs),col="red")
##boxplot(hs,col="blue")
```
```{r,echo=TRUE}
mean(hs)
median(hs)
```

These mean values are similar to the estimates in the first part of the assignment; The median values are now the same as well
The impact of imputing missing data on the estimates of the total daily number of steps was that mean and the median are now equal

Are there differences in activity patterns between weekdays and weekends?

Panel plot containing a time series plot of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). 

```{r,echo=FALSE}
hss<-subset(basedatasubstNA,select=c(stepssubstNA,date,interval))

## time series plot of avg steps per day vs interval
hssi<-melt(hss,id=c("date","interval"), measure.vars=c("stepssubstNA"))
hssii<-cast(hssi, interval~date, sum)
hcolnamessi<-c(names(hssii))
gs<-mutate(hssii,rowsmm=rowSums(hssii))
hssiii<-subset(g,select=c(interval,rowsmm))
gs1<-mutate(hssiii,avgstepsperintervalperday=hssiii$rowsmm/(length(hcolnamesi)-1))

maxintvs<-subset(gs1,gs1$avgstepsperintervalperday==max(gs1$avgstepsperintervalperday),select=c(interval))
maxintvs$interval

plotdatas<-gs1
ds0<-data.frame(intvl=c(gs$interval),AvgStepsPerIntvlPerDay=gs1$avgstepsperintervalperday)
dsx<-ds0["intvl"]
dsy<-ds0["AvgStepsPerIntvlPerDay"]
xaxisname<-"Intervals"
yaxisname<-"Avg Steps per day"
##plot(as.numeric(dx[,1]),as.numeric(dy[,1]), type="l",xaxt="n",main = "Avg Steps Per Intv Per Day(Adjusted for NA)",xlab=xaxisname,ylab=yaxisname,ylim=c(min(dy[,1]),max(dy[,1])))

plot(as.numeric(dsx[,1]),as.numeric(dsy[,1]), type="l",main = "Avg Steps Per Interval Per Day(Adjusted)",xlab=xaxisname,ylab=yaxisname,ylim=c(min(dy[,1]),max(dy[,1])))
##axis(1,1:length(dx$intvl),labels=as.char(dx$intvl))

##############################################################################################
for(x in 1:length(dsx[,1]))
{
  if(dsy[x,1]==max(dsy[,1]))
  {
    ymax<-dsy[x,1]
    xmax<-x   
    xintval<-dsx[x,1]
  }
  
}
abline(v=xintval,col="magenta")

########################################################################################
## add weekday name

wkday<-weekdays(as.Date(basedatasubstNA$date))
                
wkdytest<-vector()

for (j in 1:nrow(basedatasubstNA))
     {
       
       if (wkday[j]=="Monday") hh<-"weekday"               
       else
         if (wkday[j]=="Tuesday") hh<-"weekday"
         else
          if (wkday[j]=="Wednesday") hh<-"weekday"  
          else
            if (wkday[j]=="Thursday") hh<-"weekday"  
            else
             if (wkday[j]=="Friday") hh<-"weekday"  
              else
               if (wkday[j]=="Saturday") hh<-"weekendday"  
                 else
                  if (wkday[j]=="Sunday") hh<-"weekendday"  
       wkdytest[j]<-hh      
      }
kk<-mutate(basedatasubstNA,weekdayfactor=wkdytest)        

## panel plot weekday vs weekend
hw<-subset(kk,select=c(stepssubstNA,interval,weekdayfactor))


## time series plot of avg steps per day vs interval
hwssi<-melt(hw,id=c("weekdayfactor","interval"), measure.vars=c("stepssubstNA")) ## summarize by weekday and weekend
hwssii<-cast(hwssi, interval~weekdayfactor, mean) ## sum all steps by interval and weekdayfactor


hawswdy<-tapply(hwssii$weekday,hwssii$interval,mean)
t<-data.frame()
t<-data.frame(cbind(names(hawswdy),hawswdy))
tt<-cbind(t,"weekday")
colnames(tt)<-c("interval","avgsteps","weekdayfactor")

hawswndy<-tapply(hwssii$weekendday,hwssii$interval,mean)
tk<-data.frame()
tk<-data.frame(cbind(names(hawswndy),hawswndy))
ttkk<-cbind(tk,"weekend")
colnames(ttkk)<-c("interval","avgsteps","weekdayfactor")

gw<-rbind(tt,ttkk)

y<-as.numeric(gw$avgsteps)
avgsteps<-y
x<-as.numeric(gw$interval)
interval<-x
f<-factor(gw$weekdayfactor)


xyplot(avgsteps~interval|f,layout=c(1,2),ylab="Average Steps",xlab="Interval",type="l")
```




